### 2022-10-13 AISTATS2023/paper.tex
latest

\subsection{Case 2: Bin-Splitting}
\label{sec:simulation-examples-2}

In this simulation, we illustrate the advantages of automatic
bin-splitting against the fixed-size alternative. For this reason, we
generate a very big dataset applying dense sampling (\(N=10^6\)) and
we treat the estimation with dense fixed-size bins (\(K=10^3\)) as the
ground-truth UALE. Afterwards, we generate fewer samples (\(N=500\))
and we compare the fixed-size estimation (for several \(K\)) against
UALE automatic bin-splitting. In all set-ups, we sample from
\(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
\(\mathcal{Z^*} = \{z^*_0, \cdots,
z^*_K\}\) the sequence obtained by automatic
bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size
splitting with \(K\) bins. The evaluation metrics we report are the
average number of independent runs \(t = 30\), using each time
\(N=500\) different samples.

\paragraph{Metrics}

Evaluation is perform in terms of the Mean Absolute Error (MAE) of the
estimation of \(\mu\) and \(\sigma\) accross bins, i.e.,

% The evaluation is done with regard to two metrics counting the mean
% error per bin, where the bin error is definde as the absolute
% difference of the approximation from the ground-truth. The first
% metric, Eq.~\eqref{eq:eval_met_1}, counts the mean bin error wrt
% average effect and the second, Eq.~\eqref{eq:eval_met_2} the mean bin
% error wrt uncertainty:

@@

The ground truth UALE is the average of the dense fixed-size bins that
are within in the interval \([z_{k-1}, z_k]\). For better
interpretation of \(\mathcal{L}^{\sigma}\), we also provide the mean
error term
\(\mathcal{L}^{\rho} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
  \mathcal{Z}} \mathcal{E}(z_{k-1}, z_k) \).

@@

In this set-up, \(f(\mathbf{x}) = a_1x_1 + x_1x_2\) is a
piecewise-linear function with \(5\) different-width regions, i.e.,
\(a_1\) equals to \(\{2, -2, 5, -10, 0.5\}\) in the intervals defined
by the sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\).

@@

% We observe that in
% this case all intervals are multiples of \(0.05\). As we observe, in
% Figure~\ref{fig:ex-synth-2-1}, UALE's approximation is better than all
% fixed-size alternatives (for all \(K\)), in both mean effect
% \(\mathcal{L}^{\mu}\) and uncertainty \(\mathcal{L}^{\sigma}\). For
% understanding the importance of variable-size splitting, we analyze
% the fixed-bin error. In the top-right of
% Figure~\ref{fig:ex-synth-2-1}, we observe a positive trend between
% \(\mathcal{L}^{\mu}\) and \(K\), because as \(K\) becomes larger fewer
% points contribute to the bin estimations, leading to a worse
% approximation. The interpretation of \(\mathcal{L}^{\sigma}\) is more
% complex. For \(K\) that are multiples of \(20\), i.e.
% \(K=\{20, 40, 60, 80, 100\}\), the mean error term
% \(\mathcal{L}^{\rho}\) is low as we see in the bottom-right of
% Figure~\ref{fig:ex-synth-2-2}. In these cases, we also observe in the
% bottom-left figure, that \(L\) is increasing with \(K\) due to fewer
% points in each bin. For all other values of \(K\), the \(L\) is mainly
% affected by the mean error term \(\mathcal{E}\).

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.pdf}
  \caption{Figure 1}
  \label{fig:ex-synth-2-1}
\end{figure}

@@

In this set-up, \(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), so the
effect is non-linear in \(0 \leq x_1 \leq 1\). Due to this non
linearity, wide bins increase \(\mathcal{L}^{\rho}\) making the
uncertainty approximation biased. On the other hand, narrow bins lead
to worse approximation due to the limited number of
samples. Interestingly, in Figure~\ref{fig:ex-synth-2-2}, we observe
that UALE's bin splitting manages to compromise these conflicting
objectives.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_4.pdf}
  \caption{Figure 1}
  \label{fig:ex-synth-2-2}
\end{figure}

@@

We exclude instances with missing and outlier values. As outlier we
define the feature values which ar over three standard deviations away
from the mean feature value. We also normalize all features to
zero-mean and unit standard deviation. We split the dataset into
\(N_{tr} = 15639\) training and \(N_{test} = 3910\) test examples
(80/20 split) and we fit a Neural Network with 3 hidden layers of 256,
128 and 36 units respectively. After 15 epochs using the Adam
optimizer with learning rate \(\eta = 0.02\), the model achieves a MAE
of \(37 \cdot 10^3\) dollars.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_ale_auto.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_pdp_ice.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_5_ale_auto.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_5_pdp_ice.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_ale_auto.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_pdp_ice.pdf}\\
  \caption{Figure 1}
  \label{fig:ex-real-1}
\end{figure}

@@

We evaluate the robustness of UALE approximation. Following the
evaluation framework of Section~\ref{sec:simulation-examples-2}, we
treat as ground-truth the effects computed on the full training-set
\(N=20000\) with dense fixed-size bin-splitting (\(K=80\)). Given the
big number of samples, we make the hypothesis that the approximation
with dense binning is close to the ground truth. Afterwards, we
randomly select fewer samples, \(N=1000\), and we compare UALE
approximation with all possible fixed-size approximation. We repeat
this process \(t=30\) times. In Figure~\ref{fig:ex-real-2}, we
illustrate the mean values of
\(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\) across the 30
repetitions. We observe that UALE achieves (near) optimal
approximation in all cases. \todo{add explanation for near-constant
  feature effect vs the first case that the plot }

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_1.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_1.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_5.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_5.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_7.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_7.pdf}\\
  \caption{Figure 1}
  \label{fig:ex-real-2}
\end{figure}

>>>>

\subsection{UALE vs Fixed-Size Approximation}
\label{sec:simulation-examples-2}

In this simulation, we focuse on illustrating the advantages of UALE's
variable size bin-splitting against fixed-size bin-splitting.

\paragraph{Example set-up.}
The purpose of this experiment is two-fold. First, we show that UALE
automatically provides an approximation that is as accurate (or even
more accurate) than the best fixed-size approximation, among all
possible \(K\). Second, we illustrate that UALE favors the creation of
wide intervals when they are not adding cumulative bin-error
\(\mathcal{E}^2_{\mathcal{Z}}\). Wide intervals facilitate the
interpretation of the average effect and its uncertainty.

For measuring the accuracy of the approximation, we generate a very
big dataset with dense sampling (\(N=10^6\)) and we use dense
fixed-size bins (\(K=10^3\)). We treat this estimation as the
ground-truth UALE. Afterwards, we generate fewer samples (\(N=500\))
and we compare the fixed-size estimation (for several \(K\)) against
UALE's. We sample from \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We, also, denote as
\(\mathcal{Z^*} = \{z^*_0, \cdots, z^*_K\}\) the sequence obtained by
automatic bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the
fixed-size splitting with \(K\) bins. The evaluation is based on the
Mean Absolute Error (MAE) in estimating the bin effect \(\mu\) and the
uncertainty \(\sigma\) accross bins, i.e.,

@@

We obtaint the ground truth bin-effect, \(\mu(z_{k-1}, z_k)\), and the
ground truth uncertainty, \(\sigma(z_{k-1}, z_k)\), using the average
of the dense fixed-size bins that are within in the interval
\([z_{k-1}, z_k]\). We also provide
\(\mathcal{L}^{\rho} = \frac{1}{|\mathcal{Z}|} \sum_{k \in
  \mathcal{Z}} \mathcal{E}(z_{k-1}, z_k) \) for facilitating the
interpretation of the uncertainty error, and we call it mean residual error.

We compare UALE vs fixed-size approximation, when (a) \(f\) is
piecewise-linear and (b) \(f\) is non-linear. In all examples, we
execute \(t = 30\) independent runs, using each time \(N=500\)
different samples, and we report the mean values of the metrics.

@@

Here, the black-box function is \(f(\mathbf{x}) = a_1x_1 + x_1x_2\),
with \(5\) different-width regions, i.e., \(a_1\) equals to
\(\{2, -2, 5, -10, 0.5\}\) in the intervals defined by the sequence
\(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). We are interested in the effect of
\(x_1\) which is \(f^{\mathcal{GT}}_{\mu}(x_1) = a_1x_1\) in the
regions defined above and the uncertainty is
\(f^{\mathcal{GT}}_{\sigma}(x_1) = \sigma_2 x_1\).

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.pdf}
  \caption{Bin-Splitting, piecewise-linear function: UALE's approximation
    (Top-Left). UALE vs fixed-size approximations in terms of:
    \(\mathcal{L}^{\mu}\) (Top-Right), \(\mathcal{L}^{\sigma}\)
    (Bottom-Left), \(\mathcal{L}^{\rho}\) (Bottom-Right).}
  \label{fig:ex-synth-2-1}
\end{figure}

@@

Here, the black-box function is
\(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), so the effect is
\(f^{\mathcal{GT}}_{\mu}(x_1) = 4 x_1^2\) and the uncertainty
\(f^{\mathcal{GT}}_{\sigma}(x_1) = \sigma_2 x_1\). Due to the
non-linear effect, wide bins increase the mean residual error,
\(\mathcal{L}^{\rho}\) making the uncertainty approximation
biased. This can be clearly observed in in the bottom-right of
Figure~\ref{fig:ex-synth-2-2}. On the other hand, narrow bins lead to
worse approximation due to the limited number of
samples. Interestingly, UALE's bin splitting manages to compromise
these conflicting objectives and provide an accurate estimation, as we
see in the bottom-left of Figure~\ref{fig:ex-synth-2-2}.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_2/fig_4.pdf}
  \caption{Bin-Splitting, non-linear function: UALE's approximation
    (Top-Left). UALE vs fixed-size approximations in terms of:
    \(\mathcal{L}^{\mu}\) (Top-Right), \(\mathcal{L}^{\sigma}\)
    (Bottom-Left), \(\mathcal{L}^{\rho}\) (Bottom-Right).}
  \label{fig:ex-synth-2-2}
\end{figure}

@@

We exclude instances with missing or outlier values and we normalize
all features to zero-mean and unit standard deviation. We split the
dataset into \(N_{tr} = 15639\) training and \(N_{test} = 3910\) test
examples (80/20 split) and we fit a Neural Network with 3 hidden
layers of 256, 128 and 36 units respectively. After 15 epochs using
the Adam optimizer with learning rate \(\eta = 0.02\), the model
achieves a MAE of \(37 \cdot 10^3\) dollars.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_ale_auto.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_1_pdp_ice.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_5_ale_auto.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_5_pdp_ice.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_ale_auto.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/feature_7_pdp_ice.pdf}\\
  \caption{Real-dataset, UALE (left column) vs PDP-ICE (right column)
    for features \(x_2, x_6, x_8\) (top to bottom).}
  \label{fig:ex-real-1}
\end{figure}

@@

We evaluate the robustness of UALE approximation, following the
evaluation framework of Section~\ref{sec:simulation-examples-2}. We
treat as ground-truth the effects computed on the full training-set
\(N=20000\) with dense fixed-size bin-splitting (\(K=80\)). Given the
big number of samples, we make the hypothesis that the approximation
with dense binning is close to the ground truth. Afterwards, we
randomly select fewer samples, \(N=1000\), and we compare UALE
approximation with all possible fixed-size approximation. We repeat
this process for \(t=30\) independent runs and we report the mean
values for \(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma}\) accross the
repetitions. In Figure~\ref{fig:ex-real-2}, we observe that UALE
achieves accurate approximations in all cases; the MAE of the
uncertainty \(\mathcal{L}_\sigma\) is (near) the best of all
fixed-size approximations.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_1.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_1.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_5.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_5.pdf}\\
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_mu_err_feature_7.pdf}
  \includegraphics[width=.23\textwidth]{real_dataset_3/compare_var_err_feature_7.pdf}\\
  \caption{Real-dataset, UALE vs fixed-size approximation in terms of
    \(\mathcal{L}^{\mu}\) (left column), \(\mathcal{L}^{\sigma}\)
    (right column) for features \(x_2, x_6, x_8\) (top to bottom).}
  \label{fig:ex-real-2}
\end{figure}
